Relationships in the extremes and their influence on competition and coexistence

### 2021-09-21 paper/Paper.Rnw
editing intro

The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events, i.e., heat-waves, floods, hurricanes, fires, etc. 

@@

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 
%***DAN: Do it like you have it but shorter.
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and extent \citep{Meehl_2004, Ummenhofer_2017}. 
Thus, the rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. 
%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here... 
Changes in means can be attributed to extreme events; however changes in means are more-so a result of gradual shifts over time, whereas increased variance is more likely a product of extreme events as they push environmental variables to their extremes.
Currently, effects on competition and coexistence due to climate change are more widely researched through studying changes in the mean of environmental variables, thus less likely to be a result of extreme events. 
For example, decreased mean growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversely, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as mean rainfall levels are forecasted to decrease.    
%...and variance...
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model.  We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
%Lotetry model intro
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
lottery model, and below (in the next paragraph), we describe a hypothetical example. Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. We simulated two different communites up to \Sexpr{M} time steps to hopefully make it clear to the reader that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog}). The only distinction between the two simulations is that one depends on asymmterically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverablility when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

@@

We here pursue goal G1 from the Introduction, to augment the MTC to show how ATA
plays a role. For pedgogical clarity, we develop each step of the theory in general and for 
a simple example model, the lottery model. 
The lottery model is as follows. 
%***DAN: Note the inclusion of the model here means we will have to edit the "Model" section of sup mat, since we don't need to intro the model twice. Do later.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.

@@

>>>>

The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 

@@

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%***DAN: Possible topic sentence like thing after talking to Jasmin on 9/17: 
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 
%***DAN: Do it like you have it but shorter.
Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied, the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, have been studied in several systems.  
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance, both of which can be attributed to extreme events \citep{Hansen_2012}.
Currently, effects on competition and coexistence due to climate change are more widely researched through studying changes in the mean of environmental variables. 
Increased variance is less studied. 
% thus less likely to be a result of extreme events. 
%influenced by extreme climatic events as they increase in frequency, intensity, duration, and extent \citep{Meehl_2004, Ummenhofer_2017}. 
%Thus, the rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. 
%Jasmin tried to split this para up into effects on coexistence of changes in mean, starting here... 
Changes in means can be attributed to extreme events; however changes in means are more-so a result of gradual shifts over time, whereas increased variance is more likely a product of extreme events as they push environmental variables to their extremes.

For example, decreased mean growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversely, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as mean rainfall levels are forecasted to decrease.    
%...and variance...
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 
%...and extreme events (though this blends together with changes in variance)
Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.
%DAN: So what jasmin was trying to do was summarize a lot of lit on effects of climate change on coexistence, in these three parts
%(means, varinace, and extreme events).

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the reader that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog}). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

@@

We here pursue goal G1 from the Introduction, to augment the MTC to show how ATA
plays a role. For pedgogical clarity, we develop each step of the theory in general and for 
a simple example model, the lottery model (see \ref{section:introduction}).

%***DAN: Note the inclusion of the model here means we will have to edit the "Model" section of sup mat, since we don't need to intro the model twice. Do later.

@@

Although environmental fluctuations is helpful for coexistence and storage effect, unusual extreme fluctuations can be a hinderance on coexistence by introducing or changing ATAs between environmental variables.
